Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-04-04 writeups/Su_IGW_break.tex
responded to half of ref comments

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

@@

@@

\begin{abstract}
    In binaries composed of early-type stars and white dwarfs, the dominant
    tidal process involves the excitation of internal gravity waves (IGWs),
    which propagate towards the stellar surface, and their dissipation via
    nonlinear wave breaking. We perform 2D hydrodynamical simulations of this
    wave breaking process in a stratified, isothermal atmosphere. We find that,
    after an initial transient phase, the dissipation of the IGWs naturally
    generates a sharp critical layer, separating the lower stationary region
    (with no mean flow) and the upper ``synchronized'' region (with the mean
    flow velocity equal to the horizontal wave phase speed). While the critical
    layer is steepened by absorption of these waves, it is simultaneously
    broadened by Kelvin-Helmholtz instabilities such that, in steady state, the
    critical layer width is determined by the Richardson criterion. We study the
    absorption and reflection of incident waves off the critical layer and
    provide analytical formulae describing its long-term evolution. The result
    of this study is important for characterizing the evolution of tidally
    heated white dwarfs and other binary stars.
\end{abstract}

@@

We consider a stratified, incompressible, isothermal fluid representing a
stellar envelope or atmosphere. We approximate the equation of state as
baratropic, i.e.\ $P(\rho, T) = P(\rho)$. We study dynamics in 2D, so that fluid
variables depend only on the Cartesian coordinates $x$ and $z$. While it is
well-known wave breaking is a 3D process \citep{klostermeyer,winters1994}, the
dynamical effect of the breaking process is likely to be similar in 2D
\citep{barker_ogilvie}. We approximate the gravitational field as uniform,
pointing in the $(-\uv{z})$ direction. The plane-parallel approximation is
justified since wave breaking generally occurs near the stellar surface. The
background density stratification is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We denote background quantities
with overbars and perturbation quantities with primes.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $-g\uv{z}$. For this isothermal background, hydrostatic
equilibrium implies $P = \overline{\rho} g H$. We assume there is initially no
background flow, so $\bm{u} = \bm{u}'$. Physically, this assumption corresponds
to a non-rotating star.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}. The
solution in this linear regime is given by \citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
and is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The phase and group velocities are given by:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \p{2H}^{-2}},\\
            \bm{c}_{g} &= N\frac{\s{k_{z}^2 + \p{2H}^{-2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\s{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\s{\p{k_{z}H}^{-2}}
        \approx 0$. In the Boussinesq approximation where terms of order
        $\mathcal{O}\p{H^{-2}}$ are ignored, the phase and group velocities are
        exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

@@

Consider now the long-term evolution of the critical layer due to continuous
horizontal momentum transfer by IGWs. Any incident horizontal momentum flux
$F_a(t)$ absorbed by the fluid must manifest as additional horizontal momentum
of the shear flow. Since $\overline{U}$ cannot exceed $\overline{U}_c$, the
critical layer will instead propagate downward in response to the incident
momentum flux. The horizontal momentum of the shear flow satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - F_a(t) = 0.
\end{equation}
Assuming $\overline{U}(z > z_c) \approx \overline{U}_c$ and $\overline{U}(z <
z_c) \approx 0$, this condition becomes
\begin{equation}
    -\overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = F_a(t).\label{eq:zc_anal}
\end{equation}
If $F_a$ is constant in time, the height of the critical layer $z_c(t)$ has
analytical solution:
\begin{equation}
    z_c(t) = -H\ln \s{\exp\p{-\frac{z_c(t = 0)}{H}} +
        \frac{tF_a}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

@@

We solve Eqs.~\eqref{eq:nl_incomp},~\eqref{eq:nl_upsilon_u},
and~\eqref{eq:vol_drive} in a Cartesian box with size $L_x, L_z$. We choose
periodic boundary conditions in both the $x$ and $z$ direction. To mimic the
absence of physical boundaries at the top/bottom of the simulation domain, we damp
perturbations to zero near the top/bottom using damping zones
(see Section~\ref{ss:damping}). We expand all variables as Fourier series
with $N_x$ and $N_z$ modes, and use the $3/2$ dealiasing rule to avoid aliasing
errors in the nonlinear terms \citep{boyd}.

The geometry of our simulation domain is fixed by one further parameter, $z_0$,
the forcing location. We choose $L_z = 12.5H$ ($z$ runs from $-H$ to $11.5H$),
and the lower and upper damping zones are active for $z < 0.3H$ and $z > 9.5H$
respectively. The forcing (see Eq.~\eqref{eq:vol_drive}) is at $z_0 = 2H$ with
width $\sigma = 0.078H$, sufficiently far from the lower damping zone and
permitting sufficient room for the upward propagating wave to grow as $\propto
e^{z/2H}$. Finally, we want similar grid spacing in the $x$ and $z$ directions
(i.e.\ $L_x / N_x \sim L_z /N_z$), guided by the intuition that turbulence
generated by wave breaking is approximately isotropic, so we use $L_x = 4H$ and
$N_z / N_x = 4$.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.3H$ and $z_T = 9.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. We choose the transition width $\Delta z = 0.25H$ and damping
time $\tau = 1 / (15N)$. This prescription is similar to \citet{lecoanet_damp}
and has the advantage of being smooth, important for spectral methods. Further
details of our implementation of the fluid equations in Dedalus are described in
Appendix~\ref{se:strat_impl}.

@@

We expect the waves to follow the analytical solution given by
Eq.~\eqref{eq:uz_lin} and the corresponding $u_x(x, z, t)$; we denote this
solution $\bm{u}_{al}(x, z, t)$. The amplitude of the observed IGW in the
simulation field $\bm{u}$ relative to analytical solution $\bm{u}_{al}$ over
some region $z \in [z_b, z_t]$ can be estimated from
\begin{equation}
    A_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{al}}^2\;\mathrm{d}x\mathrm{d}z}.
        \label{eq:ahat_def}
\end{equation}
The subscript $i$ denotes the incident wave. If $\bm{u} = \bm{u}_{al}$, then
$A_i(t) = 1$. The normalization in Eq.~\eqref{eq:ahat_def} is chosen such that
the overlap between $\bm{u}, \bm{u}_{al} \propto e^{z/2H}$ is evenly weighted
throughout the integration region.

@@

% GOT HERE

@@

To perform simulations of wave breaking phenomena, we use the same setup as
described in Section~\ref{s:numerics} and Section~\ref{s:weak_sim} except for
different values of $C$ and $\nu$. In particular, we choose $C$ such that
$\abs{\xi_z k_z} = 0.1$ in the forcing zone ($z = z_0$). The linear solution
predicts $\abs{\xi_z k_z} \sim 4.25$ at the upper damping zone $z_T$. We choose
the viscosity $\nu$ so that $\mathrm{Re}$ is as large as possible across the
various resolutions. A table of our simulations can be found in
Table~\ref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Spectral resolutions and Reynolds numbers of simulations of wave
    breaking.}\label{tab:params}
\end{table}

@@

% convert yubo_000054.png -crop 2000x2000+150+250 out.png
In Fig.~\ref{fig:snapshots}, we present snapshots of $u_x$ and $\Upsilon$ at
various phases of the simulation. The flow evolves through several distinct
stages:
\begin{enumerate}
    \item At early times (top left panel), the flow resembles a linear IGW lower
        in the simulation domain but breaks down into smaller-scale features at
        higher $z$. Some characteristic swirling motion can be seen in the
        advected scalar $\Upsilon$, indicating Kelvin-Helmholtz instabilities.

@@

    \item At the end of the simulation (bottom right panel), the flow shows very
        few significant qualitative differences from the previous snapshot,
        suggesting that the latter phase of the simulation has reached a steady
        state.
\end{enumerate}

@@

The formation of the critical layer is associated with a strong shear flow. What
is the width of this layer? Inspection of Fig.~\ref{fig:snapshots} suggests the
presents of the Kelvin-Helmholtz Instability (KHI) in the critical layer. In a
stratified medium, KHI occurs when the Richardson number (Eq.~\eqref{eq:ri_def})
satisfies $\mathrm{Ri} \lesssim 1/4$ \citep[e.g.][]{shu1991physics}. It is
natural to suspect that the shear flow cannot steepen further than the onset of
KHI\@. To test this, we compute the local $\mathrm{Ri}$ for the shear flow
around the critical layer.

It is difficult to accurately measure the Richardson number, as it depends on
the derivative of the velocity. We measure $\mathrm{Ri}$ as follows: we first
assign an $\mathrm{Ri}_x(x, t)$ for every $x$ in the critical layer, then take
the median of $\mathrm{Ri}$ for the entire layer. $\mathrm{Ri}_x$ is computed
using the vertical distance over which the local $u_x$ increases from
$0.3\bar{U}_c$ to $\bar{U}_c$ (see Eq.~\eqref{eq:u_crit}). The value $0.3$ is
necessary to exclude the small mean flow generated in the weakly nonlinear
regime far below the
critical layer. This procedure can be written:
\begin{align}
    z_{CL, \min}(x, t) &\equiv \min
        \z{z\mid u_x(x, z, t) > 0.3\overline{U}_c},\\
    z_{CL, \max}(x, t) &\equiv \max
        \z{z\mid u_x(x, z, t) < \overline{U}_c},\\
    \mathrm{Ri}_x(x, t) &\equiv
        \p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}},\\
    \mathrm{Ri}(t) &\equiv \med_x\mathrm{Ri}\p{x, t}.\label{eq:ri_med_def}
\end{align}
We use the background buoyancy frequency to compute $\mathrm{Ri}$, as
fluctuations do not change $N^2$ significantly ($\sim 1\%$). To understand the
variation in $\mathrm{Ri}$ over $x$, we also compute $\min\limits_x
\mathrm{Ri}_x(x, t)$ (the maximum is very noisy). Both are shown in
Fig.~\ref{fig:nl_f_ri}. Absorbtion of incident IGWs quickly decreases the
Richardson number to between $0.25$ and $0.5$, characteristic of the onset of
the KHI\@.

>>>>

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae,
        aecompl,
        graphicx,
        booktabs,
        xcolor,
        soul,
    }
    \usepackage[T1]{fontenc}
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[caption=false]{subfig} % subfigures

@@

    \colorlet{Corr}{red}

@@

\begin{abstract}
    In binaries composed of \textcolor{Corr}{either early-type stars or white
    dwarfs}, the dominant tidal process involves the excitation of internal
    gravity waves (IGWs), which propagate towards the stellar surface, and their
    dissipation via nonlinear wave breaking. We perform 2D hydrodynamical
    simulations of this wave breaking process in a stratified, isothermal
    atmosphere. We find that, after an initial transient phase, the dissipation
    of the IGWs naturally generates a sharp critical layer, separating the lower
    stationary region (with no mean flow) and the upper ``synchronized'' region
    (with the mean flow velocity equal to the horizontal wave phase speed).
    While the critical layer is steepened by absorption of these waves, it is
    simultaneously broadened by Kelvin-Helmholtz instabilities such that, in
    steady state, the critical layer width is determined by the Richardson
    criterion. We study the absorption and reflection of incident waves off the
    critical layer and provide analytical formulae describing its long-term
    evolution. The result of this study is important for characterizing the
    evolution of tidally heated white dwarfs and other binary stars.
\end{abstract}

@@

We consider a stratified, incompressible, isothermal fluid representing a
stellar envelope or atmosphere. We approximate the equation of state as
\textcolor{Corr}{barotropic}, i.e.\ $P(\rho, T) = P(\rho)$. We study dynamics in
2D, so that fluid variables depend only on the Cartesian coordinates $x$ and
$z$. \textcolor{Corr}{While it is well known that waves break differently in two
versus three dimensions} \citep{klostermeyer,winters1994}, the dynamical effect
of the breaking process is likely to be similar in 2D \citep{barker_ogilvie}. We
approximate the gravitational field as uniform, pointing in the $(-\uv{z})$
direction. The plane-parallel approximation is justified since wave breaking
generally occurs near the stellar surface. The background density stratification
is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density. We denote background quantities
with overbars and perturbation quantities with primes.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. The constant gravitational
acceleration is $-g\uv{z}$. For this isothermal background, hydrostatic
equilibrium implies \textcolor{Corr}{$\overline{P}(z) = \overline{\rho}(z) g
H$}. We assume there is initially no background flow, so $\bm{u} = \bm{u}'$.
Physically, this assumption corresponds to a non-rotating star.

@@

In the small perturbation limit, we may linearize Eq.~\eqref{se:nl_upsilon}.
\textcolor{Corr}{These equations admit the canonical IGW solution}, given by
\citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency}, $N$, is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
and is constant. Other dynamical quantities are simply related to $u'_z$.

@@

    \item The phase and group velocities are given by:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \p{2H}^{-2}},\\
            \bm{c}_{g} &= N\frac{\s{k_{z}^2 + \p{2H}^{-2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\s{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}^{3/2}}.\label{eq:vg}
        \end{align}
        \textcolor{Corr}{The additional $\p{2H}^{-2}$ term in the denominator
        accounts for the growing amplitude of the IGW in the $z$ direction (the
        $z$ wavenumber is effectively $k_z - i / (2H)$).}
        We note $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\s{\p{k_{z}H}^{-2}}
        \approx 0$. In the Boussinesq approximation where terms of order
        $\mathcal{O}\p{H^{-2}}$ are ignored, the phase and group velocities are
        exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

@@

Consider now the long-term evolution of the critical layer due to continuous
horizontal momentum transfer by IGWs. \textcolor{Corr}{Any incident horizontal
momentum flux absorbed by the fluid, denoted $F_a(t)$, must manifest as
additional horizontal momentum of the shear flow}.
\textcolor{Corr}{Additionally, as the mean flow $\overline{U}$ cannot grow
efficiently above $\overline{U}_c$ (due to the breakdown of the linear
solution), we assume $\overline{U}$ saturates at $\overline{U}_c$, which holds
to good accuracy (see Fig.~\ref{fig:nl_fluxes}). Instead, the critical layer
must propagate downward in response to the incident momentum flux.} The
horizontal momentum of the shear flow satisfies
\begin{equation}
    \pd{}{t}\int\limits \overline{\rho}(z) \overline{U}(z, t)\;\mathrm{d}z
        - F_a(t) = 0.
\end{equation}
Assuming $\overline{U}(z > z_c) \approx \overline{U}_c$ and $\overline{U}(z <
z_c) \approx 0$, this condition becomes
\begin{equation}
    -\overline{\rho}(z_c) \overline{U}_c\rd{z_c}{t} = F_a(t).\label{eq:zc_anal}
\end{equation}
If $F_a$ is constant in time, the height of the critical layer $z_c(t)$ has
analytical solution:
\begin{equation}
    z_c(t) = -H\ln \s{\exp\p{-\frac{z_c(t = 0)}{H}} +
        \frac{tF_a}{\overline{U}_c H\overline{\rho}_0}},\label{eq:zc_sol}
\end{equation}
where $z_c(t = 0)$ is the initial critical layer height.

@@

We solve Eqs.~\eqref{eq:nl_incomp},~\eqref{eq:nl_upsilon_u},
and~\eqref{eq:vol_drive} in a Cartesian box with size $L_x, L_z$. We choose
periodic boundary conditions in both the $x$ and $z$ direction. To mimic the
absence of physical boundaries at the top/bottom of the simulation domain, we
damp perturbations to zero near the top/bottom using damping zones (see
Section~\ref{ss:damping}). We expand all variables as Fourier series with $N_x$
and $N_z$ modes, and use the $3/2$ dealiasing rule to avoid aliasing errors in
the nonlinear terms \citep{boyd}. We choose $L_z = 12.5H$ ($z$ runs from $-H$ to
$11.5H$), and the lower and upper damping zones are active for $z < 0.3H$ and $z
> 9.5H$ respectively. The forcing (see Eq.~\eqref{eq:vol_drive}) is centered at
$z_0 = 2H$ with width $\sigma = 0.078H$, sufficiently far from the lower damping
zone and permitting sufficient room for the upward propagating wave to grow as
$\propto e^{z/2H}$. Finally, we want similar grid spacing in the $x$ and $z$
directions (i.e.\ $L_x / N_x \sim L_z /N_z$), guided by the intuition that
turbulence generated by wave breaking is approximately isotropic, so we use $L_x
= 4H$ and $N_z / N_x = 4$.

@@

We aim to damp disturbances that reach the vertical boundaries of the simulation
domain without inducing nonphysical reflection. To do so, we replace material
derivatives in Eq.~\eqref{se:nl_upsilon} with:
\begin{align}
    \md{}{t} &\to \md{}{t} + \Gamma(z),\\
    \Gamma(z) &= \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}},\label{eq:Gamma}
\end{align}
where $z_B = 0.3H$ and $z_T = 9.5H$ are the boundaries of the lower and upper
damping zones respectively. This damps perturbations below $z_B$ and
above $z_T$ with damping time $\tau$ and negligibly affects the dynamics between
$z_B$ and $z_T$. \textcolor{Corr}{Most importantly, horizontal momentum is still
conserved between $z_B$ and $z_T$}. We choose the transition width $\Delta z =
0.25H$ and damping time $\tau = 1 / (15N)$. This prescription is similar to
\citet{lecoanet_damp} and has the advantage of being smooth, important for
spectral methods. Further details of our implementation of the fluid equations
in Dedalus are described in Appendix~\ref{se:strat_impl}.

@@

We expect the waves to follow the analytical solution given by
Eq.~\eqref{eq:uz_lin} and the corresponding $u_x(x, z, t)$; we denote this
solution $\bm{u}_{al}(x, z, t)$. The amplitude of the observed IGW in the
simulation field $\bm{u}$ relative to analytical solution $\bm{u}_{al}$ over
some region $z \in [z_b, z_t]$ can be estimated from
\begin{equation}
    A_i(t) = \frac{\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\p{\bm{u} \cdot \bm{u}_{al}}\;\mathrm{d}x\mathrm{d}z}
        {\int\limits_{z_b}^{z_t}\int\limits_0^{L_x}
        \overline{\rho}\abs{\bm{u}_{al}}^2\;\mathrm{d}x\mathrm{d}z}.
        \label{eq:ahat_def}
\end{equation}
The subscript $i$ denotes the incident wave. If $\bm{u} = \bm{u}_{al}$, then
$A_i(t) = 1$. \textcolor{Corr}{The factor of $\bar{\rho}$ inside the integrands
in Eq.~\eqref{eq:ahat_def} corrects for the $\propto e^{z/2H}$ growth of
$\bm{u}_{al}$; without it, $A_i(t)$ would be dominated by the contribution near
$z_t$.}

@@

@@

To perform simulations of wave breaking phenomena, we use the same setup as
described in Section~\ref{s:numerics} and Section~\ref{s:weak_sim} except for
different values of $C$ and $\nu$. In particular, we choose $C$ such that
$\abs{\xi_z k_z} = 0.1$ in the forcing zone ($z = z_0$). The linear solution
predicts $\abs{\xi_z k_z} \sim 4.25$ at the upper damping zone $z_T$.
\textcolor{Corr}{We choose the viscosity $\nu$ for each resolution to be as
small as possible while still resolving the shortest spatial scales of the wave
breaking.} A table of our simulations can be found in Table~\ref{tab:params}.
\begin{table}
    \centering
    \begin{tabular}{l c c c}
        Resolution & $\mathrm{Re}$\\\bottomrule
        $1024 \times 4096$ & $2048$\\
        $768 \times 3072$ & $1024$\\
        $512 \times 2048$ & $512$\\
        $256 \times 1024$ & $341$\\
        $256 \times 1024$ & $205$\\
        $256 \times 1024$ & $146$\\
    \end{tabular}
    \caption{Spectral resolutions and Reynolds numbers of simulations of wave
    breaking.}\label{tab:params}
\end{table}

@@

% convert yubo_000054.png -crop 2000x2000+150+250 out.png
In Fig.~\ref{fig:snapshots}, we present snapshots of $u_x$ and $\Upsilon$ at
various phases of the simulation. \textcolor{Corr}{Note that $\Upsilon \lesssim
0.1$, so the density stratification does not deviate significantly from
equilibrium.} The flow evolves through several distinct stages:
\begin{enumerate}
    \item At early times (top left panel), the flow resembles a linear IGW lower
        in the simulation domain but breaks down into smaller-scale features at
        higher $z$. Some characteristic swirling motion can be seen in the
        advected scalar $\Upsilon$, indicating Kelvin-Helmholtz instabilities.

@@

    \item At the end of the simulation (bottom right panel), the
        \textcolor{Corr}{critical layer has advanced downwards, but otherwise}
        the flow shows very few significant qualitative differences from the
        previous snapshot. This suggests that the latter phase of the simulation
        has reached a steady state.

        \textcolor{Corr}{Notably, the horizontal banded structure of $u_x$ in
        the synchronized fluid is mostly unchanged (also visible in the top
        panel of Fig.~\ref{fig:nl_fluxes}), suggesting that momentum
        redistribution and mixing within the synchronized fluid are negligible.}
\end{enumerate}

@@

The formation of the critical layer is associated with a strong shear flow. What
is the width of this layer? Inspection of Fig.~\ref{fig:snapshots} suggests the
\textcolor{Corr}{presence} of the Kelvin-Helmholtz Instability (KHI) in the
critical layer. In a stratified medium, KHI occurs when the Richardson number
(Eq.~\eqref{eq:ri_def}) satisfies $\mathrm{Ri} \lesssim 1/4$
\citep[e.g.][]{shu1991physics}. It is natural to suspect that the shear flow
cannot steepen further than the onset of KHI\@. To test this, we compute the
local $\mathrm{Ri}$ for the shear flow around the critical layer.

It is difficult to accurately measure the Richardson number, as it depends on
the derivative of the velocity. We measure $\mathrm{Ri}$ as follows: we first
assign an $\mathrm{Ri}_x(x, t)$ for every $x$ in the critical layer, then take
the median of $\mathrm{Ri}$ for the entire layer. $\mathrm{Ri}_x$ is computed
using the vertical distance over which the local $u_x$ increases from
$0.3\bar{U}_c$ to $\bar{U}_c$ (see Eq.~\eqref{eq:u_crit}). The value $0.3$ is
necessary to exclude the small mean flow generated in the weakly nonlinear
regime far below the
critical layer. This procedure can be written:
\begin{align}
    z_{CL, \min}(x, t) &\equiv \min
        \z{z\mid u_x(x, z, t) > 0.3\overline{U}_c},\\
    z_{CL, \max}(x, t) &\equiv \max
        \z{z\mid u_x(x, z, t) < \overline{U}_c},\\
    \mathrm{Ri}_x(x, t) &\equiv
        \p{\frac{N^2 \p{z_{CL, \max} - z_{CL, \min}}^2}{(0.7
            \overline{U}_c)^2}},\\
    \mathrm{Ri}(t) &\equiv \med_x\mathrm{Ri}\p{x, t}.\label{eq:ri_med_def}
\end{align}
We use the background buoyancy frequency to compute $\mathrm{Ri}$, as
fluctuations do not change $N^2$ significantly ($\sim 1\%$). To understand the
variation in $\mathrm{Ri}$ over $x$, we also compute $\min\limits_x
\mathrm{Ri}_x(x, t)$ (the maximum is very noisy). Both are shown in
Fig.~\ref{fig:nl_f_ri}. \textcolor{Corr}{Absorption} of incident IGWs quickly
decreases the Richardson number to between $0.25$ and $0.5$, characteristic of
the onset of the KHI\@.
